% SCOPE: one-paragraph abstract on the MERGED arc — motivation (self-distillation
% degrades reasoning; calibration reduces overconfidence but does NOT restore
% accuracy) -> insight (calibration != metacognitive control) -> method
% (\pmishift{} + matched-base twin) -> preliminary result (6/6 held-out wins,
% single-seed, PACKAGE-level). No obsolete numbers; preliminary labeled.
\begin{abstract}
Reinforcement learning with verifiable rewards is the standard recipe for
training reasoning models, yet the resulting models handle their own
uncertainty poorly: they are overconfident, rarely notice mid-solution errors,
and the field's default remedy is to scale the model. Self-distillation, the
natural remedy, can instead \emph{degrade} reasoning by suppressing the
epistemic, hesitation-bearing tokens that mark deliberation. A calibration-first
response---training the model to report its confidence faithfully---reduces
overconfidence but, in our earlier study, did not restore accuracy: traces grew
longer and more hesitant, yet the added hesitation was never closed by a change
of solution path. We take this as evidence that \emph{calibration is necessary
but not sufficient for metacognitive control}: the signal worth rewarding is the
causal effect a metacognitive act has on the model's evidence about the answer,
not the confidence it reports. We define in-reasoning metacognition
operationally as explicit \metaopen{}\dots\metaclose{} blocks and reinforce each
block only by the evidence shift it causes. Our \pmishift{} reward freezes a
reference model, teacher-forces the gold answer and a rule-based decoy just
before and just after a block, and scores the block by how much it moves this
gold-vs-decoy log-likelihood contrast---a form of metacognitive
self-distillation---plus a reversal bonus when a block flips the preferred
answer from decoy to gold. A matched-base twin (identical data with meta content
stripped, byte-audited hyperparameters, correctness-only GRPO) isolates the
mechanism. On GSM8K, MATH500, and AIME with Qwen3-8B, the meta arm beats its
matched twin on all six held-out comparisons (preliminary, single seed): GSM8K
$+4.0$, MATH500 $+18.8$, and AIME $+13.8$~pp (avg@16, 16k tokens, bootstrap
$p<.001$), with the gain concentrating on the hardest problems ($+34.8$~pp in
the top difficulty quartile) even though meta emission is difficulty-agnostic.
Both arms are flat from 4k to 16k tokens and the base truncates far more often,
so the meta mechanism \emph{mitigates} rather than causes the long-context
degeneration. We report these as preliminary results for a multi-head reward
package; a decomposition isolating \pmishift{} from its auxiliary terms is
ongoing.
\end{abstract}
