% SCOPE: MERGED narrative arc. [motivation] self-distillation degrades reasoning
% + calibration reduced overconfidence but did NOT restore accuracy (earlier
% calibration-era finding, ONE paragraph, NO obsolete numbers) -> [insight]
% calibration != metacognitive control; reward the causal behavioral effect, not
% the confidence report -> [method] \pmishift{} (gold-vs-decoy margin shift
% across a meta span) + matched-base causal twin, positioned against AntiSD /
% reference-likelihood line -> [result] meta arm beats matched-base on held-out,
% 6/6 (PRELIMINARY, single-seed, PACKAGE-level) -> [4 contributions].
\section{Introduction}
\label{sec:intro}

Reinforcement learning with verifiable rewards (RLVR) has become the dominant
recipe for improving mathematical reasoning in large language models: GRPO and
its descendants optimize a single scalar signal---final-answer
correctness---over sampled solution traces
\citep{shao2024deepseekmath,deepseekr1,zeng2025simplerlzoo,qwen3}. This recipe is
remarkably effective at sharpening the policy toward answers the model can
already reach, but it leaves a conspicuous gap: nothing in the objective teaches
the model to \emph{handle its own uncertainty}. Trained models remain
overconfident, commit to a flawed line of attack early, and rarely notice
mid-solution that their evidence no longer supports the path they are on
\citep{huang2024llmcannot,kadavath2022language}. The field's default remedy is
scale---larger models, more RL compute, longer traces---but scaling raises the
competence frontier without changing the character of behavior \emph{at} that
frontier: wherever a model's reliable reach ends, the same silent overconfidence
reappears.

The natural remedy for a broken uncertainty signal is to repair it directly, and
here the evidence is discouraging. Self-distillation---training a model on its
own verified outputs---can \emph{degrade} reasoning rather than sharpen it,
because conditioning on a known-correct solution systematically suppresses the
epistemic, hesitation-bearing tokens (``wait,'' ``perhaps,'' ``let me check'')
that carry deliberation, collapsing traces into short, monotone paths
\citep{kim2026selfdistill,opd2026rethinking,sdzero2026}. A calibration-first
response is to make the model \emph{report} its uncertainty faithfully. In an
earlier study on hard mathematics we trained confidence as an explicit control
signal on top of a metacognitive token layer; the intervention reduced
overconfidence, but it did not restore accuracy. Traces instead grew longer and
more hesitant while the added hesitation was never closed by an actual change of
solution path---the model learned to say it was unsure without acting on it. The
lesson we carry forward is a single sentence: \emph{calibration is necessary but
not sufficient for metacognitive control}.

Two further lines of evidence explain why. \citet{gandhi2025cognitive} show that
RL amplifies cognitive behaviors---verification, backtracking, subgoal setting,
backward chaining---only when those behaviors are already present in the base
model or its priming data; RL selects among existing behaviors rather than
installing new ones. Read together with the self-distillation degradation
results, this says that current pipelines reward the \emph{form} of
self-reflective reasoning wherever it happens to appear, but never its
\emph{function}. A model can learn to emit verification-shaped text---``let me
double-check this step''---and even to attach a well-calibrated confidence to it,
while the text changes nothing about the answer, because no term in the objective
asks whether the reflection actually moved the model's evidence. What is missing
is not a better uncertainty report but training pressure that rewards a
metacognitive act by its \emph{causal effect on the model's evidence about the
answer}.

We build exactly this. First, we define in-reasoning metacognition
operationally: the model emits \metaopen{}\dots\metaclose{} blocks inside its
chain of thought, primed by SFT stages that teach where such blocks belong and
what functional content (verify, redirect) they should carry, so that
metacognitive acts can be rewarded separately from object-level reasoning.
Second, we reinforce each block only by what it accomplishes. Our \pmishift{}
reward freezes a reference model and teacher-forces the gold answer and a
rule-based decoy at two points---just before \metaopen{} and just after
\metaclose{}---yielding a pointwise-mutual-information contrast between gold and
decoy at each point; the reward is the \emph{shift} in this contrast across the
block, plus a reversal bonus when the block flips the preferred answer from decoy
to gold and a penalty for the reverse. The shift is routed onto the meta-block
tokens alone, so the meta region is rewarded exactly insofar as it moves the
model's own gold-vs-decoy evidence: the model distills its latent evidence
contrast into explicit metacognitive text, a form of metacognitive
self-distillation. Correctness remains the primary objective; \pmishift{} is an
auxiliary shaping term.

This places our reward in a line that has recently converged on
reference-answer-likelihood signals for reasoning: rewarding a trace by the
log-probability it assigns to the correct answer \citep{yu2025rlpr,zhang2025crm},
or self-distilling from the model's own verified solutions
\citep{mulgund2026rlsd}. The closest prior work is AntiSD
\citep{shen2026antisd}, which computes a \emph{per-token} conditional PMI---each
token's log-likelihood ratio with and without a verified solution as
context---diagnoses that self-distillation rewards structural and shortcut tokens
while deflating deliberation tokens, and \emph{reverses} the sign to protect
deliberation. We differ on three axes: we reward a \emph{shift} of a
gold-vs-\emph{decoy} discriminative margin across an explicit metacognition
\emph{span}, rather than a per-token imitation signal with reversed sign, and we
pair it with a matched-base causal twin. The shortcut/boilerplate risk that
AntiSD identifies is nonetheless shared---our verify-boilerplate is their
shortcut bias---which is precisely why we build in the decomposition and placebo
tests below. Third, we train a matched-base twin: identical data with meta
content stripped, byte-audited identical non-meta hyperparameters,
correctness-only GRPO---so any difference between arms isolates the meta
mechanism rather than confounded data or tuning.

Our current results are preliminary---a single seed, and the launched meta arm is
a multi-head reward package (correctness plus \pmishift{}, calibration, emission,
format, and length terms), so what we measure is the package effect against
vanilla GRPO rather than \pmishift{} in isolation; the decomposition that
isolates it is ongoing. With those caveats, the signal is consistent. On held-out
evaluation the meta arm beats its matched-base twin on all six benchmark
comparisons: GSM8K $93.9$ vs $89.9$, MATH500 $81.5$ vs $62.7$
($+18.8$~pp, $p<.001$), and AIME $18.5$ vs $4.8$ (avg@16 at 16k tokens,
$+13.8$~pp, bootstrap $p<.001$). Both arms are flat from 4k to 16k tokens---the
gain is not bought with more decoding budget---and the base truncates far more
often (roughly $73\%$ vs $50\%$ of AIME traces), so the long-context
non-termination that plagues these models is a property the meta mechanism
\emph{mitigates}, not one it causes. The effect is sharply
difficulty-dependent---saturated on easy quartiles, $+34.8$~pp on the hardest
quartile---even though meta emission itself is difficulty-agnostic
($86$--$94\%$ across quartiles), which means aggregate comparisons understate the
effect and must be difficulty-stratified.

\paragraph{Contributions.}
\begin{enumerate}
\item An \textbf{insight and its operationalization}: calibration is necessary
  but not sufficient for metacognitive control, so the signal to reward is the
  \emph{causal effect} of a metacognitive act on the model's evidence, not the
  confidence it reports. We turn this into an executable definition of
  in-reasoning metacognition---explicit \metaopen{}\dots\metaclose{} blocks
  governing when to verify and redirect.
\item \textbf{\pmishift{}}, a metacognitive self-distillation reward that scores
  each meta block by the gold-vs-decoy evidence \emph{shift} it produces across
  the span, with save/derail reversal bonuses. We position it against the
  reference-likelihood line and its closest neighbor AntiSD
  \citep{shen2026antisd}: a span shift of a decoy-contrastive margin localized to
  a metacognition region, not a per-token sign-reversed imitation signal.
\item A \textbf{matched-base causal design and analysis suite}---a meta-stripped,
  byte-audited twin plus priming-vs-reward decomposition, save/derail flip
  analysis, placebo meta content, difficulty stratification, and
  calibration---designed to attribute any accuracy gain to the meta mechanism and
  to locate \emph{where} it pays.
\item An \textbf{empirical characterization on math reasoning}
  (GSM8K/MATH500/AIME, Qwen3-8B): preliminary single-seed, package-level evidence
  that the meta arm wins all six held-out comparisons, that the effect
  concentrates on hard problems while emission does not, and a Simpson-style
  caution that aggregate meta-vs-base comparisons must be difficulty-stratified.
\end{enumerate}
