% SCOPE: 3.1 meta-token protocol + two-stage SFT priming; 3.2 \pmishift{}
% reward with exact equations (4-arm teacher-forced PMI, clip, reversal bonus,
% META_CONTENT-region routing, group centering); 3.3 matched-base twin design.
% Grounded in src/training/dcpo_pmi_shift.py, src/training/dcpo_region.py,
% src/training/verl_sdc.py (_compute_dcpo_v4_pmi_shift_rmeta),
% src/training/_decoy_utils.py.
\section{Method}
\label{sec:method}

Our method has three components: an \emph{operational protocol} that makes
metacognition an explicit, localizable object in the token stream
(Section~\ref{sec:method:protocol}); the \pmishift{} reward, which scores each
metacognitive block by the evidence shift it causally produces in the model's
own beliefs (Section~\ref{sec:method:pmishift}); and a matched-base twin that
isolates the contribution of this mechanism from every other ingredient of the
pipeline (Section~\ref{sec:method:twin}).

\subsection{The meta protocol and two-stage SFT priming}
\label{sec:method:protocol}

\paragraph{Operationalizing metacognition.}
We define in-reasoning metacognition as an explicit, delimited intervention on
the model's own solution attempt: at any point inside its reasoning trace, the
model may open a block
\begin{center}
\metaopen{} \texttt{confidence:} $\hat{p}$ \quad \texttt{assessment:} \emph{(diagnosis of the current attempt)} \quad \texttt{action:} \emph{(verify / redirect / continue)} \metaclose{}
\end{center}
where \metaopen{} and \metaclose{} are dedicated special tokens added to the
vocabulary. The block states a confidence estimate, an assessment of the
partial solution so far, and a committed action (e.g.\ verify the current
answer, or redirect to a different approach), after which the model resumes
solving. This representation is deliberately \emph{executable} rather than
descriptive: the block sits at a known position in the trace, its content is
machine-parsable, and---crucially for Section~\ref{sec:method:pmishift}---the
contexts immediately before and after it are well defined, so the block's
effect on downstream beliefs can be measured.

\paragraph{Two-stage SFT priming.}
Following the observation that RL amplifies behaviors already present in the
base policy rather than installing new ones \citep{gandhi2025cognitive}, we
prime the protocol with two supervised stages on Qwen3-8B. \textbf{SFT-1}
teaches \emph{form and placement}: it trains on reasoning traces with
well-formed \metaopen{}\dots\metaclose{} blocks embedded \emph{inside} the
chain of thought (the \texttt{v8\_meta\_inside\_strict} corpus), so the model
reliably emits syntactically valid, correctly delimited blocks mid-solution.
\textbf{SFT-2} teaches \emph{function}: it adds traces in which the meta block
is followed by a behavioral consequence---a redirect that abandons a failing
approach, or a verification that re-derives the candidate answer---so that
blocks are not inert annotations but precede an actual change in the solving
trajectory (the \texttt{rv\_functional} corpus). After priming, the RL stage
below decides \emph{which} of these behaviors to reinforce, and only where
they demonstrably help.

\subsection{\pmishift{}: a metacognitive self-distillation reward}
\label{sec:method:pmishift}

RL training uses a GRPO-family objective in which verified answer correctness
is the primary reward and \pmishift{} is an auxiliary shaping term routed
exclusively onto meta-block tokens; the launched objective additionally
carries a small set of protocol-maintenance heads and costs (calibration,
format, emission, length, truncation), enumerated in the routing paragraph
below, whose role is to keep the meta channel well-formed rather than to
improve answers. The design principle is that a meta block
should be rewarded not for existing, and not for the correctness of the
rollout it happens to sit in, but for the \emph{belief update it causes}.

\paragraph{Two-position evidence contrast.}
Consider a rollout $y$ for prompt $x$ that contains a meta block $m$
(tag-inclusive, \metaopen{}$\dots$\metaclose{}). Let $y_{<m}$ denote the
reasoning prefix strictly before \metaopen{}, and define the two evaluation
contexts
\begin{equation}
c_{\mathrm{open}} \;=\; (x,\, y_{<m}),
\qquad
c_{\mathrm{close}} \;=\; (x,\, y_{<m},\, m),
\label{eq:contexts}
\end{equation}
i.e.\ the model's state just \emph{before} reading its own meta block and just
\emph{after} it. Let $a^{\star}$ be the gold answer rendered as a boxed answer
string, and $\tilde{a}$ a \emph{decoy}: a rule-based perturbation of the gold
answer (integer/float offsets, numerator--denominator swaps, sign flips,
dropped constants), generated deterministically from the gold string and
checked for non-equivalence to $a^{\star}$ under the same
\texttt{math\_verify}-based checker used for grading. This check is
best-effort: if the checker itself raises an exception, decoy selection
degrades to string- and float-level non-equality filters, so symbolically
equivalent decoys are excluded whenever the checker succeeds rather than
unconditionally. With a \emph{frozen}
reference model $\pi_{\mathrm{ref}}$ we teacher-force both answer strings at
both contexts---four arms per rollout, batched into a single no-gradient
forward---and compute a pointwise-mutual-information--style log-odds of gold
versus decoy at each position:
\begin{align}
\mathrm{PMI}_{\mathrm{open}}
  &= \textstyle\sum_{t \in \mathcal{D}}
     \Big[ \log \pi_{\mathrm{ref}}\!\big(a^{\star}_{t} \,\big|\, c_{\mathrm{open}}, a^{\star}_{<t}\big)
         - \log \pi_{\mathrm{ref}}\!\big(\tilde{a}_{t} \,\big|\, c_{\mathrm{open}}, \tilde{a}_{<t}\big) \Big],
\label{eq:pmiopen}\\
\mathrm{PMI}_{\mathrm{close}}
  &= \textstyle\sum_{t \in \mathcal{D}}
     \Big[ \log \pi_{\mathrm{ref}}\!\big(a^{\star}_{t} \,\big|\, c_{\mathrm{close}}, a^{\star}_{<t}\big)
         - \log \pi_{\mathrm{ref}}\!\big(\tilde{a}_{t} \,\big|\, c_{\mathrm{close}}, \tilde{a}_{<t}\big) \Big],
\label{eq:pmiclose}
\end{align}
where $\mathcal{D}$ is the set of token positions at which the (length-matched)
gold and decoy tokenizations diverge; rows whose tokenizations mismatch in
length fail closed rather than being zero-padded, which would fabricate
spurious log-probabilities. $\mathrm{PMI}_{\mathrm{open}}$ is the model's
gold-vs-decoy evidence \emph{before} its own metacognitive intervention;
$\mathrm{PMI}_{\mathrm{close}}$ is the evidence \emph{after}. Their difference
\begin{equation}
\Delta \;=\; \mathrm{PMI}_{\mathrm{close}} - \mathrm{PMI}_{\mathrm{open}}
\label{eq:shift}
\end{equation}
is the meta block's marginal contribution to moving belief toward the correct
answer: everything else about the rollout (prompt, prefix, answers, reference
model) is held fixed across the two positions, so $\Delta$ isolates the effect
of the block itself.

\paragraph{Reward with clipping and asymmetric reversal bonuses.}
The per-rollout meta reward combines a clipped continuous term with an
asymmetric bonus for genuine \emph{sign reversals} of belief:
\begin{equation}
R_{\mathrm{shift}}
\;=\;
s \cdot \mathrm{clip}\!\big(\Delta,\, -c,\, c\big)
\;+\;
\begin{cases}
+\beta_{\mathrm{save}}
  & \text{if } \mathrm{PMI}_{\mathrm{open}} < -\varepsilon
    \text{ and } \mathrm{PMI}_{\mathrm{close}} > \varepsilon
    \quad \text{(\textsc{save})}\\[2pt]
-\beta_{\mathrm{derail}}
  & \text{if } \mathrm{PMI}_{\mathrm{open}} > \varepsilon
    \text{ and } \mathrm{PMI}_{\mathrm{close}} < -\varepsilon
    \quad \text{(\textsc{derail})}\\[2pt]
0 & \text{otherwise,}
\end{cases}
\label{eq:reward}
\end{equation}
with scale $s = 1$, clip $c = 2$, $\beta_{\mathrm{save}} = 1.0$, and
$\beta_{\mathrm{derail}} = 2.0$. A \textsc{save} (a decoy$\to$gold reversal)
is exactly the event of
interest---the meta block corrected a wrong-leaning belief---while a
\textsc{derail} (gold$\to$decoy: the block talked the model out of a correct
answer) is
penalized strictly harder than a save is rewarded, making harmful
metacognition more costly than useful metacognition is profitable. The margin
$\varepsilon \geq 0$ exists as a guard against marginal-jitter reward
farming: with $\varepsilon > 0$ a reversal must clear a dead band on
\emph{both} sides of zero. Our reported runs use $\varepsilon = 0$, under
which the strict inequalities exclude only exact zero-crossings; the dead
band is an ablatable knob, not an active safeguard in the results we report.
Degenerate inputs fail closed to $R_{\mathrm{shift}}=0$:
non-finite PMI values, empty meta blocks (for which
$c_{\mathrm{close}}=c_{\mathrm{open}}$ and the shift is null by construction),
and decoys that collapse onto the gold string. A further guard---a token-set
Jaccard filter denying credit to meta blocks that near-duplicate the
preceding reasoning---is implemented but disabled (threshold $1.0$) in the
reported runs. Scoring is also selective: in each rollout only the
\emph{first} closed meta block is scored; rollouts whose block is followed by
no further content (whitespace-only continuation) and rollouts whose
delimiters cannot be trusted are excluded from the member set, while
recoverable delimiter errors (swapped, duplicated, or reversed tags) are
scored after token repair.

\paragraph{Region-masked routing and group centering.}
Rewards are converted to token-level advantages by region
(Figure~\todo{method overview figure}). Each response is partitioned into an
\textsc{answer} region (all non-meta tokens), a \textsc{meta-content} region
(the opening \metaopen{} tag plus the tokens inside the block,
\emph{excluding} the confidence-number tokens), and a \textsc{conf} region
(the confidence-number tokens). The \textsc{conf} carve-out is deliberate:
dense PMI credit on the stated confidence would reward writing a confidence
that the continuation then echoes, so those tokens are routed to the
calibration head below and never receive $R_{\mathrm{shift}}$. The closing
\metaclose{} tag serves as the anchor position for the format head. Within a
GRPO group $\mathcal{G}$ of rollouts sharing a prompt, each reward head is
centered by subtracting its group mean (no variance normalization, following
Dr.GRPO~\citep{liu2025understanding}),
$\hat{A}^{(i)} = R^{(i)} - \tfrac{1}{|\mathcal{M}|}\sum_{j \in \mathcal{M}} R^{(j)}$,
where the meta head is centered only over the member set $\mathcal{M}$ of
rollouts whose shift was actually computed---centering over non-emitting
rollouts would both dilute the baseline with structural zeros and hand those
rollouts a spurious nonzero advantage. To keep heads on a common scale, the
auxiliary heads' centered advantages are rescaled to match an
exponential-moving-average of the correctness head's mean absolute advantage
(\emph{anchor normalization}); the centered, rescaled value is then broadcast
\emph{uniformly} onto every token of its own region. The two heads of primary
scientific interest are
\begin{equation}
A_{t} \;=\; w_{\mathrm{corr}}\, \hat{A}_{\mathrm{corr}} \,
\mathbb{1}[t \in \textsc{answer}]
\;+\;
w_{\mathrm{meta}}\, \hat{A}_{\mathrm{shift}} \,
\mathbb{1}[t \in \textsc{meta-content}]
\;+\;
A^{\mathrm{aux}}_{t},
\label{eq:routing}
\end{equation}
with $w_{\mathrm{corr}} = 1.0$ and $w_{\mathrm{meta}} = 0.8$; the meta weight
is linearly warmed up from $0$ to $0.8$ over the first $80$ training steps.
The term $A^{\mathrm{aux}}_{t}$ collects three protocol-maintenance heads,
routed analogously: a Brier-style calibration head on the \textsc{conf}
tokens ($w_{\mathrm{cal}} = 0.3$), a format head anchored on the closing tag
($w_{\mathrm{format}} = 0.35$; $+1$ for a well-formed closed block, $-0.2$
for drifting or discarded blocks), and a binary emission head on the first
response token ($w_{\mathrm{emit}} = 0.1$) that counters silent collapse of
the protocol. Two scalar costs complete the objective: a length cost that
subtracts $0.08 \cdot (\text{response length} / \text{max length})$ (scaled
by the same warmup) from the correctness scalar, and a $0.3$ penalty for meta
blocks left dangling open at truncation. The correctness head is therefore
$R_{\mathrm{corr}} \in \{-1, +1\}$ (verified final-answer grading with
\texttt{math\_verify}) minus the length cost; it remains the primary
objective and never flows onto meta tokens. Conversely, $R_{\mathrm{shift}}$
never flows onto answer tokens, so improving the answer cannot directly game
the meta channel nor vice versa; the maintenance heads touch only their own
anchor positions.

\paragraph{Interpretation: metacognitive self-distillation.}
Equations~\eqref{eq:pmiopen}--\eqref{eq:routing} implement a form of
self-distillation with an unusual target. The model already carries, in its
own frozen-reference likelihoods, a latent contrast between the correct answer
and a plausible perturbation of it. \pmishift{} rewards the policy for
\emph{writing meta content that surfaces this latent evidence}---text whose
inclusion in the context measurably moves the gold-vs-decoy log-odds in the
right direction. Rather than distilling a teacher's outputs into the student,
the model distills its own evidence contrast into its explicit metacognitive
channel; the reward is high precisely when the verbalized self-assessment
does causal work.

\subsection{Matched-base causal design}
\label{sec:method:twin}

Any accuracy difference between our model and an off-the-shelf RLVR baseline
confounds the meta mechanism with the SFT data distribution, prompt format,
token budget spent on meta blocks, and RL hyperparameters. We therefore train
a \emph{matched-base twin} that differs from the \pmishift{} arm only in the
meta package defined below.

\textbf{Held identical.} The twin is trained on the same problems in the same
order, with the SFT corpora \emph{meta-stripped}: every
\metaopen{}\dots\metaclose{} block is removed while the surrounding reasoning
is preserved verbatim, so both arms see the same mathematical content. All
non-meta hyperparameters---optimizer, learning-rate schedule, group size,
rollout sampling parameters, sequence budget, training steps---are identical,
enforced by a byte-level audit of the two configurations that whitelists only
the meta-specific keys: the reward-head list, the meta/calibration/format/%
emission head weights and their warmup, the length cost, the truncation
penalty, and the meta-token and region-masking switches. Both arms are
evaluated in the same jobs with the same
seeds and the same grading (Section~\ref{sec:experiments}).

\textbf{What differs.} The twin runs vanilla GRPO with the correctness reward
only: no meta tokens, no region masking, no \pmishift{} head, and none of the
protocol-maintenance machinery of Section~\ref{sec:method:pmishift}. The
treatment arm adds, as a package: the protocol and priming of
Section~\ref{sec:method:protocol}, the \pmishift{} head, the
calibration/format/emission heads, the length cost, and the truncation
penalty of Section~\ref{sec:method:pmishift}. One inherited mechanism is
explicitly \emph{not} part of the treatment: the configuration chain
registers a counterfactual rollout-split agent (which would split each GRPO
group into with-meta and meta-banned sub-arms), but that split is gated on a
different reward source (\texttt{cf\_group}) and we verified it is inactive
under the \pmishift{} source; both arms generate all rollouts through the
same single-turn rollout path.

\textbf{Why this isolates the mechanism.} Under this design the two arms share
base model, data content, compute, and optimization, so the treatment
$-$ twin contrast estimates the causal effect of the full meta package
\{meta priming $+$ \pmishift{} shaping $+$ protocol-maintenance heads and
costs\}---not of the \pmishift{} reward in isolation; the Gandhi-style
decomposition arm in
Section~\ref{sec:exp:rq2} (meta-primed SFT followed by vanilla GRPO) further splits
this package into the contribution of priming versus RL-time shaping.
The twin also guards against a subtle failure mode we document in
Section~\ref{sec:exp:rq3}: any comparison between meta-emitting and
non-emitting \emph{subsets} of a single model conditions on the endogenous
decision to emit, whereas the twin comparison is a between-arm contrast on
the full, identical evaluation set. This matters because emission and payoff
dissociate in our runs---the treatment arm emits a meta block on 86--94\% of
responses across \emph{all} difficulty quartiles (non-selective;
Section~\ref{sec:exp:rq3}), while the accuracy benefit concentrates on the
hardest problems---so within-model subset comparisons would mix this
uniform-emission behavior with the difficulty-dependent payoff, and only the
between-arm contrast cleanly estimates the effect of the meta package.
