% SCOPE: insight-driven related work. Each subsection carries ONE argument;
% prior work is cited as support for that argument, not enumerated.
% 2.1 outcome-reward RL cannot credit the epistemic step; 2.2 self-distillation
% gives a strong but mis-aimed signal (AntiSD = closest neighbor); 2.3 the field
% converged on reference-answer-likelihood but never localized it; 2.4 reporting
% doubt is not using doubt. Funnel to our span-shift + decoy + meta-block corner.
\section{Related Work}
\label{sec:related}

\subsection{Outcome-reward RL cannot credit the step that did the epistemic work}
\label{sec:related:rlvr}

Reinforcement learning from verifiable rewards trains on one bit of feedback,
whether the final answer was correct, and that single bit cannot tell which step
earned it. The reward attaches to the whole trajectory at once, so a
verification that actually flipped the outcome looks identical, to the
optimizer, to a decorative hedge that merely co-occurred with a right answer.
This recipe is now the default for mathematical reasoning, from the
group-relative policy gradient that made it practical at scale to zero-RL
replications and production post-training pipelines
\citep{shao2024deepseekmath, deepseekr1, zeng2025simplerlzoo, qwen3}. In every
case the reflective-looking behavior that appears, the re-checking and the
backtracking, is never itself rewarded. It survives only because the answers it
accompanies happen to be correct.

Process supervision moves credit down to the level of individual steps, but it
grades each step against an external label rather than against the model's own
evidence. Scoring reasoning step by step outperforms outcome-only reward on
competition math \citep{lightman2024letsverify}. Yet the score asks whether a
step is correct by an outside rubric, not whether the step changed what the
model itself believes about the answer. A step that is textbook-correct and a
step that actually resolved the model's uncertainty are treated the same. The
signal therefore still cannot single out the moment that did epistemic work.

Reinforcement learning amplifies cognitive behaviors that are already present in
the model, and it grows their shape without ever checking that they function.
Four such behaviors, verification, backtracking, subgoal setting, and backward
chaining, predict which base models improve under RL, and priming a model with
them unlocks gains even when the primed solutions carry wrong answers
\citep{gandhi2025cognitive}. This is the canonical evidence that the substrate
must be installed before RL can select for it. It also exposes the gap we
target. Priming installs the outward form of a habit, and amplification is
indiscriminate, so nothing in the loop confirms that an amplified verification
verifies anything. What is missing is a reward that pays a metacognitive step
only when that step actually moves the model's evidence about the answer.

\subsection{Self-distillation gives a strong but mis-aimed signal, and the closest neighbor}
\label{sec:related:selfdistill}

A model's own likelihoods are a powerful training signal, but they reward the
wrong tokens unless the signal is aimed with care. Descending toward a
self-teacher on top of RLVR can degrade reasoning below plain GRPO, so a model
teaching itself is not automatically safe \citep{kim2026selfdistill}. The same
likelihoods become useful once their sign is reversed, with a frozen reference
model's signal ascended rather than descended \citep{kim2026rebellious}. The
mechanism behind both poles is now pinned down. Conditioning distillation on the
gold solution suppresses the deliberation tokens, the hedging and re-derivation
that carry multi-step search, because those tokens look improbable once the
answer is already known, which shortens exactly the reasoning that hard problems
require \citep{whydegrade2026}.

A whole literature documents this same mispricing and patches it, which shows
the pathology is robust rather than incidental. Across on-policy distillation,
work independently characterizes the failure modes and prescribes off-policy
initialization, dense self-generated supervision, single-model student-teacher
schemes, consensus gating, privileged self-distillation, precomputed teacher
log-probabilities, and joint distillation with RLVR
\citep{opd2026rethinking, sdzero2026, opsd2025, gates2025, hdpo2025,
lightning_opd2026, mulgund2026rlsd}. The number of independent recoveries is the
point. When many groups converge on the same failure and the same class of fix,
the mispricing is a structural property of gold-conditioned self-imitation, not
an artifact of one setup.

AntiSD supplies the sharpest diagnosis of which tokens are mispriced, and we
occupy a different corner of the same problem \citep{shen2026antisd}. It measures,
per token, whether knowing the answer makes a generated token more likely, and
shows that privileged context inflates structural and boilerplate tokens while
deflating the deliberation tokens that drive search; its remedy reverses that
per-token sign over the whole trajectory. Our signal differs on three axes.
First, we reward a shift, a difference across a span, rather than a level at each
token. Second, we score a gold-versus-decoy discriminative margin, asking not
just whether the answer's likelihood rises but whether it rises against a
plausible wrong answer, rather than a single sign-reversed imitation term.
Third, we localize the reward to an explicit \metaopen{}\dots\metaclose{} block
and validate the installed substrate with a matched causal twin, rather than an
accuracy curve alone. The shortcut risk AntiSD names is shared: our
verify-boilerplate is their scaffolding bias, and a span-shift reward could in
principle be earned by a formulaic ``let me check'' that moves no evidence. That
shared risk is precisely why our experiments control for it rather than assume it
away.

\subsection{The field converged on reference-answer-likelihood but never localized it}
\label{sec:related:likelihood}

Verifier-free methods have independently converged on the same underlying
quantity, the reference answer's likelihood under the model. One line replaces
the external verifier with the model's own average decoding probability of the
gold answer tokens, arguing that this intrinsic probability is the model's
self-evaluation of its reasoning \citep{yu2025rlpr}. A second reaches the same
quantity from the process side, modeling each step's reward as the probability
of staying on a correct-outcome trajectory so that the summed reward equals the
log-probability of the correct answer \citep{zhang2025crm}. Two clean instances,
arriving from opposite directions, land on reference-answer-likelihood as the
training signal.

Preference and self-rewarding methods circle the same object but score whole
responses, never where the likelihood moves or against what alternative.
Contrastive pairs, self-judged iteration, self-play, and iterative preference
optimization all reshape behavior by comparing complete responses to one another
\citep{ko2024distillm2, yang2024rlcd, yuan2024selfrewarding, wu2024metarewarding,
wu2024sppo, xu2024iterativedpo, fourhabits2026wn}. None of them asks at which
point inside a trajectory the answer's likelihood changes, and none contrasts the
gold answer against a competing wrong one. We sit inside this same
reference-answer-likelihood neighborhood, and our contribution is a narrow but
honest one: a gold-versus-decoy contrast measured as a shift across an explicit
metacognition span, rather than the gold likelihood alone scored globally.

\subsection{Reporting doubt is not using doubt}
\label{sec:related:calibration}

A long line teaches a model to report its uncertainty accurately, yet a
perfectly calibrated model can correctly doubt every hard problem and still fail
all of them. Miscalibration was made measurable through the gap between stated
confidence and observed accuracy \citep{guo2017calibration}. Large models turn
out to be mostly well-calibrated when asked after the fact whether their own
answers are true \citep{kadavath2022language}, and further work trains or prompts
them to state calibrated uncertainty in words
\citep{lin2022teaching, tian2023just}, including by rewarding calibrated
confidence directly under RL \citep{baniharouni2026rewardingdoubtreinforcementlearning}.
In all of this, calibration is the terminal object. The metric pays a model for
reporting doubt accurately, not for resolving it, so accurate doubt and better
answers remain separate achievements.

The converse line asks whether a model can use its self-assessment while
reasoning, and finds that it either fails or earns credit at the wrong
granularity. Iterative self-critique can improve outputs \citep{madaan2023selfrefine},
but on reasoning, without external feedback, intrinsic self-correction does not
improve and can degrade accuracy \citep{huang2024llmcannot}. Training-based
approaches reinforce whole successful traces without crediting any act inside
them \citep{zelikman2022star}, attach reward to the outcome of a revised attempt
so that credit lives between complete solutions \citep{kumar2024score}, or fix a
next trial in an external agent loop \citep{shinn2023reflexionlanguageagentsverbal}.
In each case the credit lands at episode granularity, not at the mid-solution
moment where the model's evidence goes wrong. Our target is exactly that moment:
within a single attempt, we reward the metacognitive block only when the model's
evidence for the right answer, measured against a decoy, actually improves across
it.
